Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2019-11-11 paper/manuscript.tex
fixing latex typos

\todo[inline]{}The current Overleaf document is \url{https://www.overleaf.com/5136551543vgjqbkzpmqkp}}

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig1/Fig1_dummy.pdf}
    \caption{{\bf What is alchemistry?}
    \todo[inline, color=green!20]{JDC: What is the concept for what this figure is supposed to communicate?}
    }
    \label{fig:illustrative-alchemical-transformation}
\end{figure}

@@

\section{Statistical mechanics can tell us why alchemical free energy calculations work}
\label{sec:theory}
Why would want to run an alchemical free energy calculation and why do they work? Suppose you want to compute the binding affinity, or free energy of binding, of a ligand $L$ to a protein $P$, given by:
\begin{euqation}
P+L\leftrightharpoons PL
\end{euqation}
The equilibrium rate of binding is given by the ratio of concentrations of product $[PL]$ and reactants $[P]$, $[L]$:
\begin{equation}
 K_b^{\circ} = c^{\circ}\frac{[[PL]}{[L][P]}
\end{equation}
In order to define a reference state the, standard state is usually defined as $1 \mathrm{mol}/\mathrm{L}$ and $c^\circ$ is the standard state concentration. From the rate of binding a binding free energy is given by:
\begin{equation}
    \Delta G_{\mathrm{bind}} = -RT\ln K_b^{\circ}
\end{equation}
One way to estimate the binding rate is by computing the probability of finding the molecular system in the bound state of the unbound state, making use of partition function ratios, i.e.

\begin{equation}
    \Delta G_{\mathrm{bind}} = -RT\ln\frac{\int_\mathrm{bound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}}{\int_\mathrm{unbound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}}
\end{equation}
Or more generally in a simplified form the free energy difference between the bound state is:
\begin{equation}
\Delta f \equiv f(\mathrm{bound}) - f(\mathrm{unbound}) = -\ln\frac{Z(\mathrm{unbound})}{Z({\mathrm{bound}})},
\end{equation}
where $Z = \int_\mathrm{bound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}$ represents the configurational integral. 
How can we compute $\Delta G_{\mathrm{bind}}$ from a molecular dynamics simulation? Running a direct simulation of the binding events is too computationally costly, meaning we cannot easily estimate the configurational integrals of the bound and unbound state from the simulation. This is where alchemical free simulations come in. We can write down a thermodynamic cycle as seen in ~\ref{fig:illustrative-alchemical-transformation}, for a relative free energy simulation. Based on this cycle we can compute the relative free energy of binding between two compounds $A$ and $B$, by going along the arrow of $\Delta G_{\mathrm{bound}}$, where we introduce a second compound that is similar to the first one. Now $\Delta G_{\mathrm{bound}}$ can be expressed as a ratio of partition functions of toluene (A) and benzyl alcohol (B). 
The defining characteristic of alchemical free energy calculations is the use of a series of alchemically-modified potential functions $U(x; \lambda)$ in which an alchemical parameter $\lambda$ modulates interactions in a manner that cannot occur in real chemical systems.
One or more simulations are used to collect data from a multitude of alchemical states to compute a free energy difference between a chemical state ($\lambda_0$) and another chemical or alchemical reference state ($\lambda_1$),
\begin{eqnarray}
\Delta f &\equiv& f(\lambda_1) - f(\lambda_0) = - \ln \frac{Z(\lambda_1)}{Z(\lambda_0)} , \label{equation:dimensionless-free-energy-difference}
\end{eqnarray}
where the dimensionless free energy $f(\lambda) \equiv \beta F(\lambda)$ is given in terms of partition functions $Z(\lambda)$,
\begin{eqnarray}
Z(\lambda) &=& \int dx \, e^{-u(x; \lambda)} .
\label{equation:partition-function-definition}
\end{eqnarray}
Here, the inverse thermal energy $\beta \equiv (k_B T)^{-1}$ where $k_B$ is the Boltzmann constant and $T$ is the absolute temperature, and the \emph{reduced potential} $u(x; \lambda)$~\cite{shirts2008statistically} is generally given by a trace over thermodynamic parameters with their conjugate dynamical variables,
\begin{eqnarray}
u(x;\lambda) &\equiv& \beta \left[ U(x;\lambda) + p \, V(x) + \sum_{i=1}^N \mu_i \, N_i(x) + \cdots \right] . \label{equation:reduced-potential}
\end{eqnarray}
Here, the collection of thermodynamic and alchemical parameters $\theta \equiv \{\beta, \lambda, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
The physical transformation of interest---such as a protein-ligand association process, or transfer free energy among phases---may involve a thermodynamic cycle that requires several free energy differences $\Delta f$ to be computed in order to produce an estimate of the overall desired quantity.

\begin{figure}
    \includegraphics[width=0.95\columnwidth]{paper/figures/fig2/Fig2.pdf}
    \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host.}
    The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent.}
    \label{fig:binding-thermodynamic-cycle}
\end{figure}
\todo[inline, color=green!20]{This section needs to be expanded and give a good introduction to the theoretical background without going into too much detail.}

@@

\paragraph{Computing the timeseries for equilibration detection}
Typically, the timeseries of note $a_t$ analyzed in automated equilibration detection is the negative logarithm of the probability density sampled by the MCMC algorithm (up to an irrelevant additive constant).
For simple independent simulations that sample $x_t \sim \pi(x ; \lambda)$, this is given by the reduced potential
\begin{eqnarray}
a_t &\equiv& - \ln \pi(x_t; \lambda) + c = u(x_t; \lambda) .
\end{eqnarray}
For a Hamiltonian replica exchange simulation where the super-replica state $(X_t, S_t) \sim \Pi(X_t, S_t)$, where $X \equiv \{x_1, \ldots, x_K\}$ are the positions and $S \equiv \{s_1, \ldots, s_K}$ is the alchemical state permutation vecftor, the negative logarithm of the super-replica reduced potential is used:
\begin{eqnarray}
a_t &\equiv& - \ln \Pi(X_t, S_t) + c = \sum_{k=1}^K u(x_{kt}; s_{kt})
\end{eqnarray}
Note that the use of the effective reduced potential is not guaranteed to pick up on all slow relaxation processes that may be coupled to the alchemical free energy, but the simplicity of its computation means it is generally appropriate for most cases.

@@

\includegraphics[width=0.95\columnwidth]{figures/overlap/Overlap_plot.pdf}

>>>>

\todo[inline]{}The current Overleaf document is \url{https://www.overleaf.com/5136551543vgjqbkzpmqkp}

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig1/Fig1.pdf}
    \caption{{\bf What is alchemistry?}
    \todo[inline, color=green!20]{JDC: What is the concept for what this figure is supposed to communicate?}
    }
    \label{fig:illustrative-alchemical-transformation}
\end{figure}

@@

\section{Statistical mechanics can tell us why alchemical free energy calculations work}
\label{sec:theory}
Why would want to run an alchemical free energy calculation and why do they work? Suppose you want to compute the binding affinity, or free energy of binding, of a ligand $L$ to a protein $P$, given by:
P+L\leftrightharpoons PL
The equilibrium rate of binding is given by the ratio of concentrations of product $[PL]$ and reactants $[P]$, $[L]$:
 K_b^{\circ} = c^{\circ}\frac{[[PL]}{[L][P]}
In order to define a reference state the, standard state is usually defined as $1 \mathrm{mol}/\mathrm{L}$ and $c^\circ$ is the standard state concentration. From the rate of binding a binding free energy is given by:
    \Delta G_{\mathrm{bind}} = -RT\ln K_b^{\circ}
One way to estimate the binding rate is by computing the probability of finding the molecular system in the bound state of the unbound state, making use of partition function ratios, i.e. 
    \includegraphics[width=0.95\columnwidth]{paper/figures/fig2/Fig2.pdf}
    \caption{{\bf Illustration of $\lambda$}}
    \label{fig:what_is_lambda}
    \Delta G_{\mathrm{bind}} = -RT\ln\frac{\int_\mathrm{bound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}}{\int_\mathrm{unbound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}}
Or more generally in a simplified form the free energy difference between the bound state is:
\Delta f \equiv f(\mathrm{bound}) - f(\mathrm{unbound}) = -\ln\frac{Z(\mathrm{unbound})}{Z({\mathrm{bound}})},
where $Z = \int_\mathrm{bound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}$ represents the configurational integral. 
How can we compute $\Delta G_{\mathrm{bind}}$ from a molecular dynamics simulation? Running a direct simulation of the binding events is too computationally costly, meaning we cannot easily estimate the configurational integrals of the bound and unbound state from the simulation. This is where alchemical free simulations come in. We can write down a thermodynamic cycle as seen in ~\ref{fig:illustrative-alchemical-transformation}, for a relative free energy simulation. Based on this cycle we can compute the relative free energy of binding between two compounds $A$ and $B$, by going along the arrow of $\Delta G_{\mathrm{bound}}$, where we introduce a second compound that is similar to the first one. Now $\Delta G_{\mathrm{bound}}$ can be expressed as a ratio of partition functions of toluene (A) and benzyl alcohol (B). 
The defining characteristic of alchemical free energy calculations is the use of a series of alchemically-modified potential functions $U(x; \lambda)$ in which an alchemical parameter $\lambda$ modulates interactions in a manner that cannot occur in real chemical systems.
One or more simulations are used to collect data from a multitude of alchemical states to compute a free energy difference between a chemical state ($\lambda_0$) and another chemical or alchemical reference state ($\lambda_1$),
\begin{eqnarray}
\Delta f &\equiv& f(\lambda_1) - f(\lambda_0) = - \ln \frac{Z(\lambda_1)}{Z(\lambda_0)} , \label{equation:dimensionless-free-energy-difference}
\end{eqnarray}
where the dimensionless free energy $f(\lambda) \equiv \beta F(\lambda)$ is given in terms of partition functions $Z(\lambda)$,
\begin{eqnarray}
Z(\lambda) &=& \int dx \, e^{-u(x; \lambda)} .
\label{equation:partition-function-definition}
\end{eqnarray}
Here, the inverse thermal energy $\beta \equiv (k_B T)^{-1}$ where $k_B$ is the Boltzmann constant and $T$ is the absolute temperature, and the \emph{reduced potential} $u(x; \lambda)$~\cite{shirts2008statistically} is generally given by a trace over thermodynamic parameters with their conjugate dynamical variables,
\begin{eqnarray}
u(x;\lambda) &\equiv& \beta \left[ U(x;\lambda) + p \, V(x) + \sum_{i=1}^N \mu_i \, N_i(x) + \cdots \right] . \label{equation:reduced-potential}
\end{eqnarray}
Here, the collection of thermodynamic and alchemical parameters $\theta \equiv \{\beta, \lambda, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
The physical transformation of interest---such as a protein-ligand association process, or transfer free energy among phases---may involve a thermodynamic cycle that requires several free energy differences $\Delta f$ to be computed in order to produce an estimate of the overall desired quantity.

\begin{figure}
    \includegraphics[width=0.95\columnwidth]{paper/figures/fig3/Fig3.pdf}
    \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host.}
    The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent.}
    \label{fig:binding-thermodynamic-cycle}
\end{figure}
\todo[inline, color=green!20]{This section needs to be expanded and give a good introduction to the theoretical background without going into too much detail.}

@@

\paragraph{Computing the timeseries for equilibration detection}
Typically, the timeseries of note $a_t$ analyzed in automated equilibration detection is the negative logarithm of the probability density sampled by the MCMC algorithm (up to an irrelevant additive constant).
For simple independent simulations that sample $x_t \sim \pi(x ; \lambda)$, this is given by the reduced potential
\begin{eqnarray}
a_t &\equiv& - \ln \pi(x_t; \lambda) + c = u(x_t; \lambda) .
\end{eqnarray}
For a Hamiltonian replica exchange simulation where the super-replica state $(X_t, S_t) \sim \Pi(X_t, S_t)$, where $X \equiv \{x_1, \ldots, x_K\}$ are the positions and $S \equiv \{s_1, \ldots, s_K}$ is the alchemical state permutation vector, the negative logarithm of the super-replica reduced potential is used:
\begin{eqnarray}
a_t &\equiv& - \ln \Pi(X_t, S_t) + c = \sum_{k=1}^K u(x_{kt}; s_{kt})
\end{eqnarray}
Note that the use of the effective reduced potential is not guaranteed to pick up on all slow relaxation processes that may be coupled to the alchemical free energy, but the simplicity of its computation means it is generally appropriate for most cases.

@@

\includegraphics[width=0.95\columnwidth]{paper/figures/fig8/Fig8.pdf}
